Supplement: Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2

### 2023-08-31 go3.Rmd
clarified H and G, added bold notation

---
title: "Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2
"
shorttitle: "Scale ALDEx2"
author:
- name: Greg Gloor, Michelle Pistner Nixon, Justin Silverman
  affiliation: Dep't of Biochemistry, University of Western Ontario, Penn State
  email: ggloor@uwo.ca
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
output:
  BiocStyle::pdf_document: default
package: ALDEx2
abstract: |
    Introduction to scale simulation and FDR correction with ALDEx2.
---

@@

More formally, if the true information in the  environment being sampled can be summarized by counts of features (genes, taxa, etc) in a matrix, then the elements of the matrix can be decomposed into its composition (relative information) and its scale (total sum). We can formally state this as \(W = W^{\parallel} \times W^{\perp}\). If the matrix has \(N\) samples and \(D\) parts, then we can uniquely identify \(W_{dn}\) as the \(d\)^{th} feature in sample \(n\).  Any data that we collect by sequencing is thought to contains only proportional information and we can represent the corresponding sequenced matrix as \(Y\). Compositional data analysis makes the strong assumption  that \(Y_{n} = Y^{\parallel}_{n} = W^{\parallel}_{n}\); in other words that relative information is the only information available post-sequencing. However, the values after sequencing are only estimates of the true underlying values that we want to estimate and different technical and biological replicates vary. Note that under this strong assumption no corresponding estimate is obtained for the scale of the sample \(Y^{\perp}_{n}\).

@@

Recently, [,@nixon2023scale] showed that geometric mean of \(Y^{\parallel}_{n}\) was related to the geometric mean of \(W_n\) and so was related to \(W^{\perp}_n\). This realization opened the possibility that adding uncertainty to the geometric mean could provide a buffer of uncertainty on it that could be used to control for  the unacknowledged bias that the geometric mean was a perfect point estimate of the underlying system. The bias around  could be acknowledged by adding uncertainty when the geometric mean of \(Y^{\parallel}_{n}\)  was calculated. In terms of information theory, the geometric mean of \(Y^{\parallel}_{n}\) is strongly correlated with Shannon's Information \(I\) (supplemental figure XX), suggesting that  some knowledge of \(W\) is contained in the post-sequencing data that is not strictly compositional. This makes intuitive sense because  underlying systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim I_n\).

The realization that a strong assumption about scale is built into each denominator explains why HTS data are notoriously fickle as analysis of the same dataset with different workflows, tools and assumptions giving widely different results when benchmarked[@Thorsen:2016aa;@Quinn:2018aa;@Gardner:2019aa]. This can be understood as the result of unacknowledged bias about the scale estimate which inevitably leads to large Type 1 error rates as sample sizes increase [@Gustafson2015]. Thus, the realization that the denominator is a scale estimate opens up the possibility that we can supply alternate scale estimates and use this to determine which features are likely to be called as differentially abundant even when scale was taken into account, and those features that are differentially abundant only under strictly assumed normalizations.

The ALDEx2 R package [@fernandes:2013] represents a general purpose toolbox for Bayesian estimation of HTS datasets. ALDEx2 was designed originally to convert point estimates of \(Y^{\parallel}_{dn}\) into posterior distributions through Monte-Carlo sampling from the Dirichlet distribution, where the posterior more accurately represented variation on a per-sample basis[@gloorAJS:2016]. Each Monte-Carlo replicate is normalized using the CLR and used to calculate summary and test statistics, and finally  expected values and confidence intervals are reported across the Monte-Carlo replicates. The CLR calculation uses as the denominator the geometric mean of each Monte-Carlo replicate of \(Y^{\parallel}_{dn}\) with the geometric mean representing a point estimate of scale, i.e. of \(Y^{\perp}_{n}\). We  now include a posterior distribution of the scale into this process by sampling \(Y^{\perp}_n\) from a log-Normal distribution and using these as the denominator. The scale estimates can be calculated on a per-experiment, per-group or even a per-sample basis.

An advantage of incorporating scale is that the analysis can be made much more robust and that differences in scale can be accounted for explicitly. The examples below show how incorporating scale provides robust and interpretable differential abundance estimates in several different datasets.

@@

We start by examining the dispersion or variance of the data as counts and as the logarithm of the normalized counts as calculated by DESeq2 (log(RLE)) and by ALDEx2 (clr). The actual counts of data derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010] as seen in Panel A of Figure 1. This relationship is why most tools model the counts with the Negative Binomial distribution[@polyester:2016], and use this distribution as the basis for batch correction[@Zhang:2020ab]. However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [Robinson:2010, Love:2014aa], or on  the clr values[@fernandes:2013] both of which are  log-ratios. The mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013, @Love:2014aa].

```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,  as logarithms of counts, and as clr values. Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale. Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first. Panel C shows the equivalent values calculated by ALDEx2. The expected clr value for each transcipt are plotted vs. the expected dispersion. In panels B and C the amount of dispersion reaches a minimum at moderate  values. The red line in each panel shows the loess line of fit to the mid-point of the distributions. The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value. The dashed grey line in panels B and C represent the maximum of the lowest 10\\% dispersion." }

@@

Here we can see that while the variance or dispersion always increases with increasing raw read count, it actually decreases when measured on the log-ratio transformed data[@fernandes:2013; @Love:2014aa] and reaches a minimum at some mid-point of the distribution. This makes the counter-intuitive suggestion that genes with moderate expression have more predictable expression than genes with very high expression such as housekeeping genes. This is at odds with the known biology of cells where single cell counting of housekeeping transcripts shows that they are both highly expressed and have little intrinsic variation[@Taniguchi:2010aa]. Furthermore, the actual amount of dispersion is very small and for many transcripts is almost negligible. To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion below the dashed grey line are statistically significantly different (75% with DESeq2, 69% with ALDEx2), suggesting that low dispersion estimates lead to many false positives because of the underestimation of dispersion. The work of [,@nixon2023scale] suggests that this is because the biased estimate of \(Y^{\perp}\) that is  calculated leads to a false estimate of the scale. Thus, the actual variation of highly expressed genes is not captured accurately by current approaches to analysis and supports the idea that HTS have unacknowledged bias because scale is not taken into account.

Using either DESeq2 or ALDEx2, we observe that a majority of transcripts are statistically significantly different between groups even with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.01 ;`r length(sig.des)` or `r length(sig.ald)` of `r nrow(yst)` transcripts. Clearly, the necessary assumption of most features being invariant is not justified. Applying the rule of thumb of at least a $2^{1.4}$ fold change [@Schurch:2016aa] reduces these outputs to `r length(which(res@listData$padj < 0.01 & abs(res@listData$log2FoldChange) > 1.4))` for DESeq2 and to `r length(which(x.all$we.eBH < 0.01 & abs(x.all$diff.btw) > 1.4))` for ALDEx2.

@@

In the case of asymmetry, the use of a user-specified scale model can be very useful for otherwise difficult to analyze datasets such as meta-transcriptomes and in-vitro selection datasets where the majority of features can change. We showed one such example in Figure 3 where the dataset was highly asymmetrical, and the TMM and RLE normalizations cannot fully move all the housekeeping genes to the midline of no difference (supplemental). Incorporating differential scale on a per-group basis moves the mass of the data towards the midline of no difference and so affects both Type I and Type II error rates. Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant. In this analysis, transcripts that were previously not classed as differentially abundant are now called as significantly different, and the housekeeping transcripts move from being significantly different to not being identified as such. While we acknowledge that some prior information on which housekeeping transcripts should not be classed as DA is needed, we suggest that this information is widely available and used when performing the gold-standard quantitative PCR test of differential abundance [@Thellin:1999aa,@SEQC/MAQC-III-Consortium:2014aa]. Thus, the use of this prior knowledge is not unique to our approach.

>>>>

---
title: "Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2
"
shorttitle: "Scale ALDEx2"
author:
- name: Greg Gloor
  affiliation: Dep't of Biochemistry, University of Western Ontario
  email: ggloor@uwo.ca
- name: Michelle Pistner Nixon
  affiliation: College of Information Sciences and Technology, Pennsylvania  State University
- name: Justin Silverman
  affiliation: College of Information Sciences and Technology, Department of Statistics, Department of Medicine, Pennsylvania  State University
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
output:
  BiocStyle::pdf_document: default
package: ALDEx2
abstract: |
    Introduction to scale simulation and FDR correction with ALDEx2.
---

@@

More formally, if the true information in the  environment being sampled can be summarized by counts of features (genes, taxa, etc) in a matrix, then the elements of the matrix can be decomposed into its composition (relative information) and its scale (total sum). We can formally state this as \(\mathbf{W} = \mathbf{W}^{\parallel} \times \mathbf{W}^{\perp}\). If the matrix has \(N\) samples and \(D\) parts, then we can uniquely identify \(\mathbf{W}_{dn}\) as the \(d^{th}\) feature in sample \(n\).  Any data that we collect by sequencing is thought to contain only proportional information and we can represent the corresponding sequenced matrix as \(\mathbf{Y}\). Compositional data analysis makes the strong assumption  that \(\mathbf{Y}_{n} = \mathbf{Y}^{\parallel}_{n} = \mathbf{W}^{\parallel}_{n}\); in other words that relative information is the only information available post-sequencing. However, the values after sequencing are only estimates of the true underlying values that we want to estimate and different technical and biological replicates vary. Note that under this strong assumption no corresponding estimate or information is obtained for the scale of the sample \(\mathbf{Y}^{\perp}_{n}\).

@@

Recently, [@nixon2023scale] showed that certain normalizations implied that geometric mean \(G_n\) of \(\mathbf{Y}^{\parallel}_{n}\)  was related to the scale of \( \mathbf{W} \) (i.e., \(\mathbf{W}^{\perp}_n\)). From an information theoretic perspective, \(G_n\) is strongly correlated with Shannon's entropy, indicating that some knowledge of the complexity of \( \mathbf{W} \) is available post-sequencing (Supplemental). Recognition of the relationship between the underlying scale and the \(G_n\) led to the realization that this could introduce latent bias into analyses. The possibility of this bias has been unacknowledged until now because the geometric mean was assumed to be a perfect point estimate of the scale of the underlying system. This suggested that adding uncertainty to the geometric mean could provide a buffer of uncertainty and acknowledge the bias introduced by the normalization.

The ALDEx2 R package [@fernandes:2013] represents a general purpose toolbox for Bayesian estimation of HTS datasets. ALDEx2 was designed originally to convert point estimates of \(\mathbf{Y}^{\parallel}_{dn}\) into posterior distributions through Monte-Carlo sampling from the Dirichlet distribution, where the posterior more accurately represented variation on a per-sample basis[@gloorAJS:2016]. Each Monte-Carlo replicate is normalized using the CLR and used to calculate summary and test statistics, and finally  expected values and confidence intervals are reported across the Monte-Carlo replicates. The CLR calculation uses as the denominator the geometric mean of each Monte-Carlo replicate of \(\mathbf{Y}^{\parallel}_{dn}\) with the geometric mean representing a point estimate of scale, i.e. of \(\mathbf{Y}^{\perp}_{n}\). We  now include a posterior distribution of the scale into this process by sampling \(\mathbf{Y}^{\perp}_n\) from a log-Normal distribution and using these as the denominator. The scale estimates can be calculated on a per-experiment, per-group or even a per-sample basis.

An advantage of incorporating scale is that  analyses can be made much more robust and that differences in scale can be accounted for explicitly. The examples below show how incorporating scale provides robust and interpretable differential abundance estimates in several different datasets.

@@

We start by examining the dispersion or variance of the data as counts and as the logarithm of the normalized counts as calculated by DESeq2 (log(RLE)) and by ALDEx2 (clr). The actual counts of data derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010] as seen in Panel A of Figure 1. This relationship is why most tools model the counts with the Negative Binomial distribution[@polyester:2016], and use this distribution as the basis for batch correction[@Zhang:2020ab]. However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [@Robinson:2010; @Love:2014aa], or on  the clr values [@fernandes:2013] both of which are  log-ratios. The mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013; @Love:2014aa].

```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,  as logarithms of counts, and as clr values. Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale. Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first. Panel C shows the equivalent values calculated by ALDEx2  in which the expected clr value for each transcipt are plotted vs. the expected dispersion. The red line in each panel shows the loess line of fit to the mid-point of the distributions. In panels B and C the amount of dispersion reaches a minimum at moderate  values.  The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value. The values below the dashed grey line in panels B and C represent those below the first decile of dispersion." }

@@

Here we can see that while the variance or dispersion always increases with increasing raw read count, it actually decreases when measured on the log-ratio transformed data [@fernandes:2013; @Love:2014aa] and reaches a minimum at some mid-point of the distribution. This makes the counter-intuitive suggestion that genes with moderate expression have more predictable expression than genes with very high expression such as housekeeping genes. This is at odds with the known biology of cells where single cell counting of housekeeping transcripts shows that they are both highly expressed and have little intrinsic variation [@Taniguchi:2010aa]. Furthermore, the actual amount of dispersion is very small and for many transcripts is almost negligible. To show this point more clearly, the majority of the transcripts in the lowest decile of dispersion below the dashed grey line are statistically significantly different (75% with DESeq2, 69% with ALDEx2), suggesting that low dispersion estimates lead to many false positives because of the underestimation of dispersion. The work of [@nixon2023scale] suggests that this is because the biased estimates of \(G_n\) that are  calculated leads to a false estimate of the scale. Thus, the actual variation of highly expressed genes is not captured accurately by current approaches to analysis and supports the idea that HTS have unacknowledged bias because scale is not taken into account.

Using either DESeq2 or ALDEx2, we observe that a majority of transcripts are statistically significantly different between groups even with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.01; i.e. `r length(sig.des)` or `r length(sig.ald)` of `r nrow(yst)` transcripts are statistically significant. Clearly, the necessary assumption of most features being invariant is not justified. Applying the rule of thumb of at least a $2^{1.4}$ fold change [@Schurch:2016aa] reduces these outputs to `r length(which(res@listData$padj < 0.01 & abs(res@listData$log2FoldChange) > 1.4))` for DESeq2 and to `r length(which(x.all$we.eBH < 0.01 & abs(x.all$diff.btw) > 1.4))` for ALDEx2.

@@

In the case of asymmetry, the use of a user-specified scale model can be very useful for otherwise difficult to analyze datasets such as meta-transcriptomes and in-vitro selection datasets where the majority of features can change. We showed one such example in Figure 3 where the dataset was highly asymmetrical, and the TMM and RLE normalizations cannot fully move all the housekeeping genes to the midline of no difference (supplemental). Incorporating differential scale on a per-group basis moves the mass of the data towards the midline of no difference and so affects both Type I and Type II error rates. Differential scale has the property that if the differences in underlying scale of the system is large, then the sign of the differential scale is irrelevant. In this analysis, transcripts that were previously not classed as differentially abundant are now called as significantly different, and the housekeeping transcripts move from being significantly different to not being identified as such. While we acknowledge that some prior information on which housekeeping transcripts should not be classed as DA is needed, we suggest that this information is widely available and used when performing the gold-standard quantitative PCR test of differential abundance [@Thellin:1999aa,@SEQC/MAQC-III-Consortium:2014aa]. Furthermore, the assumption that housekeeping genes should not generally be included in differential abundance analysis is implicit in the dual p-value, fold-change cutoff approach in widespread use. Thus, the use of this prior knowledge is not unique to our approach.

### 2023-08-31 supplement_figs.Rmd
clarified H and G, added bold notation

# GM is correlated with Shannon's entropy and includes

```{r info, echo=F,warning=F, message=F,comment=F, fig.cap="Plot of Shannon's entropy (I) vs geometric mean (G) for each sample in three different datasets. The groups that each sample belong to are highlighted as filled or open circles. Each group in each dataset has different entropy with the groups in the selex and metatranscriptome datasets being highly distinct."}
# Gm ~ Im
# plotting geometric mean vs Shannon's entropy
# shows differential information by group
devtools::load_all('~/Documents/0_git/ALDEx_bioc')
data(selex)

load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.all.Rda'))
url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
yst <- read.table(url, header=T, row.names=1)

# remove the one gene with 0 reads
yst <- yst[rownames(yst) != "YOR072W-B",]
# Gierlinski:2015aa
yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL

par(mfrow=c(1,1))
plot(apply(selex+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) }),
  apply(selex+0.5, 2, function(x) mean(log2(x))), xlim=c(2.5,11), ylim=c(-0.5,10), 
  xlab="Shannon's entropy", ylab='Geometric mean', pch=c(rep(19,7),rep(1,7)),
  col=rgb(0,0,0,0.5))
points(apply(yst+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) }),apply(yst+0.5, 2, function(x) mean(log2(x))), col=rgb(1,0.6,0,0.5), pch=c(rep(19,44), rep(1, 42)))
points(apply(ko.both.all+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) }),apply(ko.both.all+0.5, 2, function(x) mean(log2(x))), col=rgb(1,0,0,0.5),
 pch=c(rep(1,8), rep(19,28), rep(1,8)))

legend(3,10, legend=c('selex', 'transcriptome','metatranscriptome', 'low entropy group', 'high entropy group'),
  col=c('darkgrey', 'orange', 'red', 'grey','grey'), pch=c(19,19,19,1,19))
```

```{r move}
load('analysis/x.s.mu.all.Rda')
load('analysis/x.all.Rda')

>>>>

# GM can correlate with Shannon's entropy

We can also think about this relationship from an information theoretic point of view. Empirically in most datasets the geometric mean of \(Y^{\parallel}_{n}\) is strongly correlated with Shannon's Information \(H\) (supplemental figure XX) suggesting that  some knowledge of \(W\) is contained in the post-sequencing data that is not strictly compositional. Indeed, the logarithm of the geometric mean \( G \) and \( H \) can be understood from an information theoretic point of view to be an unweighted \( G \) or a weighted measure \( H \) of 'surprisal' in the dataset (supplement). Intuitively,  underlying systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim H_n\).

Entropy \( H \) and the geometric mean \( G \) (or its logarithm \( \log_2 G \)) are not simple to relate algebraically, but they can be understood in terms of what they are measuring if their description is rephrased in a common language. Recall have a \(D \times N\) matrix of counts \(W\) decomposed into the proportions for the \(n^{th}\) sample \(W^{\parallel}_n\) (or the equivalent probability distribution \( p(w_n) \) ), and its scale \(W^{\perp}_n\).

For notational simplicity assume a single discrete random variable \( X \) with a probability distribution \( p(x) \) over \(1 \dots d \) features. The entropy \( H(X) \) in bits is:

\[
H(X) = -\sum_{i=1}^{d} p_i \log_2 p_i
\]

and for the same distribution log2 of the geometric mean \( G \) is:

\[
\log_2 G = \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
\]

As defined here \( H(X) \) is a weighted total of the uncertainty or 'surprisal' contained in \( p(x) \), and relates to the amount of information we would need to have in order to reproduce  \( p(x) \). Conversely, \( G \) is an unweighted average of the same distribution.  However, \( G \) is used as the denominator to calculate the centred log-ratio normalization (clr):

\[
clr = \log_2(p_i) - \log_2 G 
\]

over \(i=1 \dots d \).  This compares each \( p_i \) with the geometric mean \( G \). Thus \( G \) can be interpreted as a measure of difference for how far each \( p_i \) is from \( G \), or in information theoretic terms as an unweighted measure of the  mean surprisal for the distribution.

Thus, we can thus understand \( H(X) \) as a measure of the total weighed surprisal and  \( G \)  as a measure of the average unweighted surprisal for \( p(x) \). The total and the mean surprisal are related by the number of terms in \( p(x) \) and by the weighting factor for each term.

These two measures are expected to have different behaviours in different distributions of \( p_i \). In the case of a uniform distribution both \( H(X) \) and \( G \) are maximal since \( p(x) \) is equally and identically distributed. Thus, we expect that they are positively correlated here. In a Normal or a skewed distribution, we also anticipate a positive correlation because both are affected in the same direction by outlier values. In very sparse  datasets, the two measures could become uncoupled because \( H(X) \) could ascribe some uncertainty to the large number of low probability events, while \( G \) would tend to be very small. Here these two measures could be either uncorrelated or exhibit negative correlation. We can see this distributional behaviour in different datasets.

```{r info, echo=F,warning=F, message=F,comment=F, fig.cap="Plot of Shannon's entropy (H) vs geometric mean (G) for each sample in different datasets. The groups that each sample belong to are highlighted as filled or open circles. Each group in each dataset has different entropy with the groups in the selex and metatranscriptome datasets being highly distinct."}
# Gm ~ Im
# plotting geometric mean vs Shannon's entropy
# shows differential information by group
devtools::load_all('~/Documents/0_git/ALDEx_bioc')
data(selex)

load(url('https://raw.githubusercontent.com/ggloor/datasets/main/ko.both.all.Rda'))
url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
yst <- read.table(url, header=T, row.names=1)
url <- "https://raw.githubusercontent.com/ggloor/datasets/main/meta16S.tsv"
rRNA <- read.table(url, header=T, row.names=1, sep='\t')
url <- "https://raw.githubusercontent.com/ggloor/datasets/main/singleCell.tsv"
ss <- read.table(url, header=T, row.names=1, sep='\t')
ss <- ss[,c(1:100,1501:1601)]
# remove the one gene with 0 reads
yst <- yst[rownames(yst) != "YOR072W-B",]
# Gierlinski:2015aa
yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL

HS <- apply(selex+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
GS <- apply(selex+0.5, 2, function(x) mean(log2(x/sum(x))))

HY <- apply(yst+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
GY <- apply(yst+0.5, 2, function(x) mean(log2(x/sum(x))))

HM <- apply(ko.both.all+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
GM <- apply(ko.both.all+0.5, 2, function(x) mean(log2(x/sum(x))))

H16 <- apply(rRNA+0.5, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
G16 <- apply(rRNA+0.5, 2, function(x) mean(log2(x/sum(x))))

Hss <- apply(ss+0.1, 2, function(x) {-1 * sum( (x/sum(x)) * log2( (x/sum(x)) ) ) })
Gss <- apply(ss+0.1, 2, function(x) mean(log2(x/sum(x))))

par(mfrow=c(2,3))
plot(HS, GS, ylab="Geometric mean", xlab="Shannon's entropy",
  pch=c(rep(19,7),rep(1,7)), col=rgb(0,0,0,0.5))
title(main='selex')

plot(HY,GY, 
  ylab="Geometric mean", xlab="Shannon's entropy", pch=c(rep(19,7),rep(1,7)),
  col=rgb(1,0.6,0,0.5))
title(main='transcriptome')

plot(HM, GM, col=rgb(1,0,0,0.5), pch=c(rep(1,8), rep(19,28), rep(1,8)), 
  ylab="Geometric mean", xlab="Shannon's entropy")
title(main='meta-transcriptome')

plot(H16,G16, col=rgb(0,1,1,0.5),
  pch=c(rep(1,198), rep(19,161)), 
  ylab="Geometric mean", xlab="Shannon's entropy")
title(main='16S rRNA')

plot(Hss,Gss, col=rgb(0.5,1,0,0.5),
 pch=c(rep(1,100), rep(19,100)), 
  ylab="Geometric mean", xlab="Shannon's entropy")
title(main="single cell")

# empty plot for legend
plot(selex[,1], selex[,2], xlab=NA, ylab=NA, axes=F,xlim=c(-10,-1), ylim=c(-10,-1))
legend(-10,-1, legend=c('selex', 'transcriptome','metatranscriptome', '16S', 'SS', 'low entropy group', 'high entropy group'),
  col=c('darkgrey', 'orange', 'red', 'cyan', 'greenyellow','grey','grey'), pch=c(19,19,19,19,19,1,19))

```{r dist}
rep.Inf <- function(x){ x[x == -Inf] <- min(x[x > -Inf]) -1 }

# check correlation of H and G for different distributions
sel.prop <- apply(selex, 2, function(x) x/sum(x))
yst.prop <- apply(yst, 2, function(x) x/sum(x))

L.yst.prop <- log2(yst.prop)
L.sel.prop <- log2(sel.prop)

L.yst <- apply(L.yst.prop, 2, rep.Inf)

L.sel <- apply(L.sel.prop, 2, rep.Inf)

```
```{r move}
load('analysis/x.s.mu.all.Rda')
load('analysis/x.all.Rda')
